---
title: "Prioritizing potential aquaculture"
author: "Hylaea Miller"
date: "11/26/2025"
format: pdf
execute: 
  message: false
  warning: false
---
## Background


## Data


## Load Packages

```{r}
# Load libraries
library(here)
library(sf)
library(tidyverse)
library(tmap)
library(terra)
library(kableExtra)
```

## Import data

Import the tree datasets: Sea Surface Temperature, Bathymetry and the Exclusive Economic Zones

```{r}
# Read in raster sea surface temperature data
sea_temp_2008 <- rast(here("data", "data", "average_annual_sst_2008.tif"))
sea_temp_2009 <- rast(here("data", "data", "average_annual_sst_2009.tif"))
sea_temp_2010 <- rast(here("data", "data", "average_annual_sst_2010.tif"))
sea_temp_2011 <- rast(here("data", "data", "average_annual_sst_2011.tif"))
sea_temp_2012 <- rast(here("data", "data", "average_annual_sst_2012.tif"))

# Read in raster sea bathymetry data
bathymetry <- rast(here("data", "data", "depth.tif"))

# Read in vector economic zones data
economic_zones <- st_read(here("data", "data", "wc_regions_clean.shp"))

```
Check if the datasets have the same coordinate reference systems
```{r}
print(st_crs(bathymetry) == st_crs(sea_temp_2008)) # True
print(st_crs(bathymetry) == st_crs(economic_zones)) # True

```

## Process data

Find the mean SST from 2008-2012 (e.g. create single raster of average SST)

```{r}
sst_mean <- (sea_temp_2008 + sea_temp_2009 + sea_temp_2010 + sea_temp_2011 + sea_temp_2012) / 5
```

```{r}
crs(sst_mean) <- "EPSG:4326"
```

```{r}
plot(sst_mean)
```

Convert average SST from Kelvin to Celsius
hint: subtract by 273.15

```{r}
# Covert by subtracting 273.15
sst_mean_c <- sst_mean - 273.15
```

```{r}
plot(sst_mean_c)
```

Crop depth raster to match the extent of the SST raster
note: the resolutions of the SST and depth data do not match


```{r}
bathy_crop <- crop(bathymetry, sst_mean_c)
```


resample the depth data to match the resolution of the SST data using the nearest neighbor approach




```{r}
# Resample the bathymetry raster using method = "near"
bathymetry_resample <- resample(bathy_crop, y = sst_mean_c, method = "near")
```


```{r}
plot(bathymetry_resample)
```

Check that the depth and SST match in resolution, extent, and coordinate reference system
hint: can the rasters be stacked?

```{r}
# Check whether the resolution match using the res() function

if(all(res(bathymetry_resample) == res(sst_mean_c))){
  print("resolutions match")
} else {
  print("resolutions do not match")
}

```

```{r}
# Check whether the extents match using the ext() function

if(ext(bathymetry_resample) == ext(sst_mean_c)){
  print("extents match")
} else{
  print("extents do not match")
}
```

```{r}
# check and transform coordinate reference systems
if(crs(bathymetry_resample) == crs(sst_mean_c)) {
  print("Coordinate reference systems match")
} else{
  warning("Updating coordinate reference systems to match")
  # transform data to match
  bathymetry_resample <- project(bathymetry_resample, crs(sst_mean_c))
}
```


## Find suitable locations for Oysters

Oysters
Research has shown that oysters need the following conditions for optimal growth:

Sea surface temperature: 11-30°C
Depth: 0-70 meters below sea level

Reclassify SST and depth data into locations that are suitable for oysters
hint: set suitable values to 1 and unsuitable values to 0

```{r}
summary(bathymetry_resample)
minmax(bathymetry_resample)
global(bathymetry_resample, "range", na.rm = TRUE)
```


```{r}
# create reclassification matrix for the SST
sst_rcl_mat <- matrix(c(-Inf, 11, 0,
                    11, 30, 1,
                    30, Inf, 0),
                  ncol = 3, byrow = TRUE)

# use reclassification matrix to reclassify SST raster
sst_reclassified <- classify(sst_mean_c, rcl = sst_rcl_mat)

# create reclassification matrix for the depth
depth_rcl_mat <- matrix(c(-Inf, -70, 0,
                    -70, 0, 1,
                    0, Inf, 0),
                  ncol = 3, byrow = TRUE)


# use reclassification matrix to reclassify depth raster
depth_reclassified <- classify(bathymetry_resample, rcl = depth_rcl_mat)
```



```{r}
plot(sst_reclassified)
```
```{r}
plot(depth_reclassified)
```


Find locations that satisfy both SST and depth conditions

```{r}
# Select only the cells that multiply iqual 1
suitable_cells <- sst_reclassified * depth_reclassified
#lapp
```

```{r}
 suitable_oysters <- sst_reclassified == 1 & depth_reclassified == 1
```

```{r}
plot(suitable_oysters, main = "Oyster Suitability\n (True = Suitable, False = Unsuitable)")
```

```{r}
plot(suitable_cells)
```


 

## Determine the most suitable EEZ

Select suitable cells within West Coast EEZs


```{r}
# check and transform coordinate reference systems
if(crs(economic_zones) == crs(sst_mean_c)) {
  print("Coordinate reference systems match")
} else{
  warning("Updating coordinate reference systems to match")
  # transform data to match
  bathymetry_resample <- transform(economic_zones, crs(sst_mean_c))
}
```
Rasterize the EEZ data

```{r}
ez_rast <- rasterize(economic_zones, suitable_cells, field = "rgn_id")
```

```{r}
# Calculate cell area in km²
cell_area_km2 <- cellSize(suitable_cells, unit = "km")
```

```{r}
# Calculate suitable area (multiply suitable cells by cell area)
suitable_area <- suitable_cells * cell_area_km2

```

```{r}
# Extract suitable area values for each EEZ polygon
eez_areas <- extract(suitable_area, economic_zones, fun = "sum", na.rm = TRUE)
```

```{r}
# Add the results to the economic_zones data
economic_zones$suitable_area_km2 <- eez_areas[, 2]
```

```{r}
# Sort by suitable area (descending)
economic_zones_ranked <- economic_zones[order(-economic_zones$suitable_area_km2), ]
```



### Visualize the most suitable EEZ
```{r}

# Convert to data frame and select columns
eez_table <- as.data.frame(economic_zones_ranked)[, c("rgn", "suitable_area_km2")]

# Display ranked EEZ results in formatted table
kable(eez_table, 
      caption = "Top economic zones for Oysters",
      col.names = c("Region", "Suitable Area (km²)"),
      digits = 2, row.names = FALSE) %>% 
  kable_styling(font_size = 14,
                full_width = FALSE)

```


