---
title: "potential_aquaculture"
author: "Hylaea Miller"
date: "11/11/2025"
format: pdf
---
## Background


## Data


## Load Packages

```{r}
# Load libraries
library(here)
library(sf)
library(tidyverse)
library(tmap)
library(stars)
library(terra)
```

## Import data

Import the tree datasets: Sea Surface Temperature, Bathymetry and the Exclusive Economic Zones

```{r}
# Read in raster sea surface temperature data
sea_temp_2008 <- read_stars(here("data", "data", "average_annual_sst_2008.tif"))
sea_temp_2009 <- read_stars(here("data", "data", "average_annual_sst_2009.tif"))
sea_temp_2010 <- read_stars(here("data", "data", "average_annual_sst_2010.tif"))
sea_temp_2011 <- read_stars(here("data", "data", "average_annual_sst_2011.tif"))
sea_temp_2012 <- read_stars(here("data", "data", "average_annual_sst_2012.tif"))

# Read in raster sea bathymetry data
bathymetry <- read_stars(here("data", "data", "depth.tif"))

# Read in vector economic zones data
economic_zones <- st_read(here("data", "data", "wc_regions_clean.shp"))

```
Check if the datasets have the same coordinate reference systems
```{r}
print(st_crs(bathymetry) == st_crs(sea_temp_2008)) # True
print(st_crs(bathymetry) == st_crs(economic_zones)) # True

```
## Process data

Find the mean SST from 2008-2012 (e.g. create single raster of average SST)

```{r}
sst_mean <- (sea_temp_2008 + sea_temp_2009 + sea_temp_2010 + sea_temp_2011 + sea_temp_2012) / 5
```

```{r}
plot(sst_mean)
```

Convert average SST from Kelvin to Celsius
hint: subtract by 273.15
```{r}
sst_mean_c <- sst_mean - 273.15
```

```{r}
plot(sst_mean_c)
```

Crop depth raster to match the extent of the SST raster
note: the resolutions of the SST and depth data do not match
resample the depth data to match the resolution of the SST data using the nearest neighbor approach




```{r}
# Convert the star object into raster to resample 
bathymetry_rast <- rast(bathymetry)
sst_rast <- rast(sst_mean_c)
```

Resample the bathymetry raster 

```{r}

bathymetry_resample <- resample(bathymetry_rast, y = sst_rast, method = "bilinear")
```

Crop the bathymetry raster to match extent of the SST
```{r}
bathy_resample_crop <- crop(bathymetry_resample, sst_rast)
```


```{r}
plot(bathy_resample_crop)
```

Check that the depth and SST match in resolution, extent, and coordinate reference system
hint: can the rasters be stacked?

```{r}
# Check whether the extents match using the ext() function
if(ext(bathy_resample_crop) == ext(sst_rast)){
  print("extents match")
} else{
  print("extents do not match")
}
```

## Find suitable locations

Oysters
Research has shown that oysters need the following conditions for optimal growth:

sea surface temperature: 11-30°C
depth: 0-70 meters below sea level

Reclassify SST and depth data into locations that are suitable for oysters
hint: set suitable values to 1 and unsuitable values to 0
```{r}
print(minmax(sst_rast))
print(minmax(bathy_resample_crop))
```


```{r}
# create reclassification matrix for the SST
sst_rcl <- matrix(c(-Inf, 11, 0,
                    11, 30, 1,
                    30, Inf, 0),
                  ncol = 3, byrow = TRUE)

# use reclassification matrix to reclassify SST raster
sst_reclassified <- classify(sst_rast, rcl = sst_rcl)

# create reclassification matrix for the depth
depth_rcl <- matrix(c(-Inf, 0, 0,
                    0, 70, 1,
                    70, Inf, 0),
                  ncol = 3, byrow = TRUE,)


# use reclassification matrix to reclassify depth raster
depth_reclassified <- classify(bathy_resample_crop, rcl = depth_rcl)
```

```{r}
plot(sst_reclassified)
```
```{r}
plot(depth_reclassified) + plot(suitable_oysters)
```

Find locations that satisfy both SST and depth conditions
```{r}
# check and transform coordinate reference systems
if(crs(sst_reclassified) == crs(depth_reclassified)) {
  print("Coordinate reference systems match")
} else{
  warning("Updating coordinate reference systems to match")
  # transform data to match
  depth_reclassified <- project(depth_reclassified, sst_reclassified)
}
```
```{r}
crs(sst_reclassified) == crs(depth_reclassified)
```

```{r}
suitable_oysters <- sst_reclassified == 1 & depth_reclassified == 1
```


```{r}
plot(suitable_oysters)
```

```{r}
# check whether the disaggregated version matches the original
if(identical(depth_reclassified, suitable_oysters)){
  print("disaggregated data matches original")
} else {
  warning("disaggregated data does not match original")
}
```

## Determine the most suitable EEZ

Select suitable cells within West Coast EEZs

```{r}

```

find area of grid cells

```{r}

```


find the total suitable area within each EEZ
hint: it might be helpful to rasterize the EEZ data

```{r}

```

